---
title: "Initial data exploration"
author: "Andrew Ba Tran"
date: 2018-05-27T21:13:14-05:00
categories: ["R"]
tags: ["R", "projects", "workflow"]
weight: 4
slug: data-structures
---

## Vectors

A **vector** is a sequence of data elements of the same basic type. The parts that consist of a vecctor are called **components** or **elements**.

```{r vectors1}
vec1 <- c(1,4,6,8,10)
vec1
```

A vector `vec` is explicity constructed by the concatenation function `c()`.

```{r vectors2}
vec1[5]
```

Elements in vectors can be addressed by standard `[i]` indexing

```{r vectors3}
vec1[3] <- 12
vec1
```

One of the elements in the array is replaced with a new number. 

```{r vectors4}
vec2 <- seq(from=0, to=1, by=0.25)
vec2
```

This shows another useful way of creating a vector: the `seq()` or sequence function.

```{r vectors5}
sum(vec1)
```

Some calculations. 
```{r vectors6}
vec1 + vec2
```

If you add up two vectors of the same length, the first elements of both vectors are summed, and the second elements, etc., leading to a new vector length of 5.

## Matrices

**Matrices** are two-dimensional vectors.

It looks like this

```{r matrix}
mat <- matrix(data=c(9,2,3,4,5,6), ncol=3)
mat
```

The argument `data` specifies which numbers should be in the matrix.

Use either `ncol` to specify the number of columns or `nrow` to specify the number of rows.

Matrix-operations are similar to vector operations

```{r matrix2}
mat[1,2]
```

Elements of a matrix can be addressed in the usual way

```{r matrix3}
mat[2,1]
```

When you want to select a whole row, you leave the spot for the column number empty and vice versa for columns.

```{r matrix4}
mean(mat)
```

This is how a function would work with a matrix as an argument.

## Data frames

If you're used to working with spreadsheets, then *data frames* will make the most sense to you in R.

It's a matrix with names above the columns for headers. 

This means you can call and use one of the columns without knowing in which position it is.

```{r df1}
t <- data.frame(x=c(11,12,14), y=c(19,20,21), z=c(10,9,7))
t
```

A typical data frame built from arrays. The columns have the names `x`, `y`, and `z`.

```{r df2}
mean(t$z)
```

Instead of using `mean(t[,3])` like you would with a matrix, you can select the column `z` from the `t` data frame with the `$` sign.


```{r df3}
mean(t[["z"]])
```

Here's an alternative way to refer to the `z` column of the `t` data frame. But you will rarely use this method.

## Lists

Another basic structure in R is a *list*. 

The main advantage of lists is that the "columns" they're not really ordered in columns any more, but are more of a collectoin of vectors) don't have to be of the same length, unlike matrices and data frames.

Kind of like json files are structured.

```{r list1}
L <- list(one=1, two=c(1,2), five=seq(0,1, length=5))
L
```

This is how a list would appear in the workspace

```{r list2}
names(L)
```

How to find out what's in the list

```{r list3}
L$five + 10
```

How to refer to and use the numbers in the example list

Functions for working with objects


	length(x)
	dim(x)
	ncol(x)
	nrow(x)
	str(x)
	summary(x)
	View(x)
	rm(x)
	save(x, file="myfilename.rdata")
	load(file="myfilename.rdata")


Data structures

scalar / array / matrix / array / dataframe / list

Vectors

One dimensional arrays

a <- c(1, 2, 5, 3, 6, -2, 4)
b <- c("one", "two", "three")
c <- c(TRUE, TRUE, TRUE, FALSE, TRUE, FALSE)

Vectors(2)

Identifying elements

a <- c(1, 2, 5, 3, 6, -2, 4)
a[3]
[1] 5
a[c(1, 3, 5)]
[1] 1 5 6
a[2:6] 
[1] 2  5  3  6 -2

Data frame

	Rectangular array of data
	More general than a matrix - different columns can contain different modes of data (numeric, character, etc.)
	Similar to datasets in SAS, SPSS, and Stata

mydata <- data.frame( col1, col2, …, coln)

Creating a data frame

patientID <- c(111, 208, 113, 408)
age <- c(25, 34, 28, 52)
sex <- c(1,2,1,1)
diabetes <- c("Type1", "Type2", "Type1", "Type1")
status <- c(1,2,3,1)

patientdata <- data.frame(patientID, age, sex, diabetes, status)

patientdata

Specifying elements of a data frame

patientdata[1:2]

patientdata[c(diabetes", "status")]

patientdata$age

patientdata[1:2]

patientdata[c(1,3),1:2]

patientdata[2:3, 1:2]

Factors

	Data structure specifying categorical (nominal) or ordered categorical (ordinal) variables

	Tells R how to handle that variable in analyses


	Very important and misunderstood
Any variable that is categorical or ordinal should usually be stored as a factor.

patientdata$sex <- factor(patient$sex, levels=c(1, 2),
labels=c("Male", "Female"))

associates 1=Male, 2=Female
Treats sex as a categorical variable in all analyses What happens to sex=5?

patientdata$status <- factor(status, 
ordered=TRUE,
levels=c(1, 2, 3),
labels=c("Poor", "Improved", "Excellent"))

associates 1=Poor, 2=Improved, 3=Excellent Treats status as an ordinal variable in all analyses

WHEN DO FACTORS MATTER?
important for statistical analysis
Well, it determines order of categories in charts

List example

g <- "My First List"
h <- c(25, 26, 18, 39)
j <- matrix(1:10, nrow = 5) k <- c("one", "two", "three")
mylist <- list(title = g, ages = h, j, 
k)   

mylist[[2]] 
[1] 25 26 18 39
mylist[["ages"]][[1] 25 26 18 39

Formats for R



