---
title: "Handling strings"
description: "Strategies on dealing with strings in data"
author: "Andrew Ba Tran"
date: 2018-05-27T21:13:14-05:00
categories: ["R"]
tags: ["R", "wrangling", "stringr"]
weight: 4
slug: strings
---



<div style="position: relative; padding-bottom: 56.25%; height: 0; overflow: hidden;">
  <iframe src="//www.youtube.com/embed/ApMM6cEEoiQ?t=3s" style="position: absolute; top: 0; left: 0; width: 100%; height: 100%; border:0; encrypted-media" allowfullscreen title="YouTube Video"></iframe>
</div>

We're going to use the **stringr** package to manipulate text.

```{r install, warning=F, message=F}
#install.packages("stringr")
library(stringr)

library(dplyr)
```

Each function starts with `str_`

Let's load this data in:

```{r import, warning=F, message=F}
messy <- data.frame(name=c("Bill Smith", "jane doe", "John Forest-William"),
                    email=c("bsmith@themail.com", "jdoe@themail.com", "jfwilliams$geemail.com"),
                    income=c("$90,000", "$140,000", "E8500"),
                    phone=c("(203) 847-334", "207-999-1122", "2128487345"),
                    activites=c("fishing, sailing, planting flowers", "reading, raising flowers, biking", "hiking, fishing"))

messy
```

What problems do you see?

**Tasks**

1. Split name into First name and Last name
2. Convert names to title case
3. Create a new variable identifying bad email addresses
4. Convert income to a new number in US dollars
5. Create a new variable containing area code
6. Creating a new variable counting how many activities each person is engaged in
7. Break activities into a set of useful dummy codes

| Order of elements in date-time | Parse function |
| ------ | ------------------------ |
| `str_length()` | figure out length of string |
| `str_c()` | combine strings |
| `str_sub()`    | substitute string |
| `str_detect()`    | detect string in string |
| `str_match()`    | does string match |
| `str_count()`    | count strings |
| `str_split()`   | split strings |
| `str_to_upper()`   | convert string to upper case |
| `str_to_lower()`   | convert string to lower case |
| `str_to_title()`   | convert the first letter of each word to upper case |
| `str_trim()`   | eliminate trailing white space |


## String length

`str_length(string)` counts the number of characters in each element of a string or character vector.

```{r str_length}
x <- c("Bill", "Bob", "William")
str_length(x)
```

## Combine strings

`str_c(strings, sep="")`

It's like the equivalent of =concatenate in Excel.

But there are a couple of quirks

```{r str_c}
data <- data.frame(place=c("HQ", "HQ", "HQ"),
                   id=c("A", "B", "C"),
                   number=c("001", "002", "003"))

data
```

We can add a string to each value in the *number* column this way:

```{r str_c2}
data <- data %>% 
  mutate(combined=str_c("Num: ", number))

data
```

```
# A couple options that would've done the same thing:

data$combined <- str_c("Num: ", data$number)

# or 

data <- data %>% 
  mutate(combined=str_c("Num", number, sep=": "))
```

You can also pass the variable `collapse` to `str_c()` if you're turning an array of strings into one.

```{r str_c3}
data <- data.frame(place=c("HQ", "HQ", "HQ"),
                   id=c("A", "B", "C"),
                   number=c("001", "002", "003"))

data

data %>% 
  group_by(place) %>% 
  summarize(ids_combined=str_c(number, collapse="-"))
```

## subset strings

`str_sub(strings, start, end)` extracts and replaces substrings

```{r str_sub}
x <- "Dr. James"

str_sub(x, 1, 3)

```

```{r str_sub2}
str_sub(x, 1, 3) <- "Mr."
x
```

Negative numbers count from the right.

```{r str_sub3}
x <- "baby"
str_sub(x, -3, -1)
str_sub(x, -1, -1) <- "ies"
```


## detect matches

`str_detect(strings, pattern)` returns T/F

```{r str_detect1}
x <- c("Bill", "Bob", "David.Williams")
x
str_detect(x, "il")
```

## count matches

`str_count(strings, pattern)` count number of matches in a string

```{r str_count}
x <- c("Assault/Robbery/Kidnapping")
x
str_count(x, "/")

# How many offenses
str_count(x, "/") + 1
```

## extract matches

```{r str_extract}
x <- c("bsmith@microsoft.com", "jdoe@google.com", "jfwilliams@google.com")
str_extract(x, "@.+\\.com$")
```

## split strings

`str_split(string, pattern)` split a string into pieces

```{r str_split}
x <- c("john smith", "mary todd", "bill holis")

str_split(x, " ", simplify=TRUE)

first <- str_split(x, " ", simplify=TRUE)[,1]
last  <- str_split(x, " ", simplify=TRUE)[,2]
```


## replace a pattern

`str_replace(strings, pattern, replacement)` replace a pattern in a string with another string

```{r str_replace}
x <- c("john smith", "mary todd", "bill holis")
str_replace(x, "[aeiou]", "-")

str_replace_all(x, "[aeiou]", "-")
```

## change case

`str_to_upper(strings)` is upper case
`str_to_lower(strings)` is lower case
`str_to_title(strings)` is title case

```{r x_case}
x <- c("john smith", "Mary Todd", "BILL HOLLIS")

str_to_upper(x)
str_to_lower(x)
str_to_title(x)
```

## trim strings

`str_trim(strings)` remove white space at the beginning and end of string

```{r str_trim}
x <- c(" Assault", "Burglary ", " Kidnapping ")
str_trim(x)
```


----

## Your turn

Challenge yourself with [these exercises](http://code.r-journalism.com/chapter-3/#section-handling-strings) so you'll retain the knowledge of this section.

Instructions on how to run the exercise app are on the [intro page](https://learn.r-journalism.com/en/wrangling/) to this section.




-----

<span style="color:gray">© Copyright 2018, Andrew Ba Tran</span>

