---
title: "Dealing with dates"
description: "Introduction on how to deal with dates"
author: "Andrew Ba Tran"
date: 2018-05-27T21:13:14-05:00
categories: ["R"]
tags: ["R", "wrangling", "dates", "lubridate"]
weight: 5
slug: dates
---


<div style="position: relative; padding-bottom: 56.25%; height: 0; overflow: hidden;">
  <iframe src="//www.youtube.com/embed/yt9x6PNYvlw?t=3s" style="position: absolute; top: 0; left: 0; width: 100%; height: 100%; border:0; encrypted-media" allowfullscreen title="YouTube Video"></iframe>
</div>

Dates come in as characters, most of the time.

You'll need to convert them into a date variable

We'll be using the **lubridate** package.

Here's an example of a **character variable** that might be in a data frame.

```{r ex1}
some_date <- "12-31-1999"
```

Convert that date into a **date variable** with the function **mdy()**

```{r ex2, warning=F, message=F}
# If you don't have lubridate installed yet uncomment the line below and run it
#install.packages("lubridate")

# NOTE: IF YOU GET AN ERROR ABOUTZ NOT HAVING A PACKAGE CALLED stringi
# UNCOMMENT AND RUN THE LINES BELOW IF YOU HAVE A WINDOWS MACHINE

#install.packages("glue", type="win.binary")
#install.packages("stringi", type="win.binary")
#install.packages("stringr", type="win.binary")
#install.packages("lubridate", type="win.binary")

# UNCOMMENT AND RUN THE LINES BELOW IF YOU HAVE A MAC MACHINE

#install.packages("glue", type="mac.binary")
#install.packages("stringi", type="mac.binary")
#install.packages("stringr", type="mac.binary")
#install.packages("lubridate", type="mac.binary")

library(lubridate)

mdy(some_date)
```

The `mdy()` function is very versatile. It stand for month-date-year.

And it'll be able to parse any version of that (with slashes or commas, or dashes) as long as that's the order of the information. 

Check it out:

```{r ex3}

data <- data.frame(First=c("Charlie", "Lucy", "Peppermint"),
                   Last=c("Brown", "van Pelt", "Patty"),
                   birthday=c("10-31-06", "2/4/2007", "June 1, 2005"))

data$DOB <- mdy(data$birthday)

data

```

## Reading dates

| Order of elements in date-time | Parse function |
| ------ | ------------------------ |
| year, month, day | `ymd()` |
| year, day, month | `ydm()` |
| month, day, year    | `mdy()` |
| day, month, year    | `dmy()` |
| hour, minute    | `hm()` |
| hour, minute, second    | `hms()` |
| year, month, day, hour, minute, second    | `ymd_hms()` |

## Accessing date parts

| Date component | Function |
| ------ | ------------------------ |
| Year | `year()` |
| Month | `month()` |
| Week    | `week()` |
| Day of year  | `yday()` |
| Day of month  | `mday()` |
| Day of week | `wday()` |
| Hour  | `hour()` |
| Minute  | `minute()` |
| Second  | `ymd_hms()` |
| Time zone  | `ymd_hms()` |

Now that we have the date in the right format, we can extract data from it with the functions above.

```{r example_more}
data$year <- year(data$DOB)
data$month <- month(data$DOB, label=TRUE)
data$day <- day(data$DOB)
data$weekday <- wday(data$DOB, label=TRUE, abbr=FALSE)

data
```

## Date arithmetic

The function `difftime()` extracts the number of days between two dates that are passed to it

```{r math}
# We're going to use the now() function which brings in the date for today

today <- now()
data$age <- difftime(today, data$DOB)

data
```

And how does that translate into years? 

With some math. We'll have to turn the column into a number, first.

```{r math2}
data$age_years <- as.numeric(data$age) / 365.25 #.25 because of leap years

data
```

That's a pretty good start for now. To see more functions and examples, check out [the vignette](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html) for **lubridate**.



----

## Your turn

Challenge yourself with [these exercises](http://code.r-journalism.com/chapter-3/#section-dealing-with-dates) so you'll retain the knowledge of this section.

Instructions on how to run the exercise app are on the [intro page](https://learn.r-journalism.com/en/wrangling/) to this section.



-----

<span style="color:gray">© Copyright 2018, Andrew Ba Tran</span>


